---
title: "__Status of COVID-19 cases in Ontario by Public Health Unit (PHU)__"
author: "Jeanice () | Rahma Nur Jannati (1009715499) | Vika Permata Joe (1010309649)"
date: "2024-04-05"
output:
  pdf_document: default
  html_document: default
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
data <- read_csv("cases_by_status_and_phu.csv")
```

# A. Description of The Variables and The Data

## a) Data Description

The dataset provides information on the status of COVID-19 cases in Ontario, including the number of active cases, resolved cases, and deaths, broken down by Public Health Unit. It serves as a valuable resource for monitoring the spread of COVID-19 and assessing the effectiveness of public health interventions and strategies. The dataset covers a specific time period with daily snapshots of COVID-19 testing and cases reported across various Public Health Units in Ontario. It includes both current and archived data on COVID-19 cases.

However, only data recorded after March 11, 2022, are included in the analysis of this report. This decision was made due to the change of the methodology used to count COVID-19 deaths by excluding deaths not caused by COVID-19, thereby ensuring a more accurate representation of the impact of the virus. As a result of the methodological change, the columns "Resolved_Cases" and "Deaths" represent data recorded after March 11, 2022, while the columns "Archived_Resolved_Cases" and "Archived_Deaths" represent data posted publicly prior to the methodological change.

## b) Variable Description

* File_date: The date of the reported data.
* PHU_Name: The name of the Public Health Unit (PHU) in Ontario.
* PHU_Num: A unique identifier for each Public Health Unit.
* Active_Cases: Number of active COVID-19 cases.
* Resolved_Cases: Number of resolved COVID-19 cases after March 11, 2022 (excluding deaths not caused by COVID).
* Deaths: Number of deaths due to COVID-19 after March 11, 2022.
* Archived_Resolved_Cases: Archived number of resolved COVID-19 cases posted publicly prior to the methodological change. This variable is not used in this report.
* Archived_Deaths: Archived number of deaths due to COVID-19 posted publicly prior to the methodological change. This variable is not used in this report.

# B. Background about The data

The data is collected by Public Health Units (PHUs) in Ontario and The Ontario Ministry of Health with the aim of tracking COVID-19 cases across different public health units in Ontario.

# C. Overall Research Question

The main research question guiding our analysis of the dataset is:
How have the COVID-19 testing and case trends evolved across Public Health Units (PHUs) in Ontario since March 11, 2022, and what factors are associated with variations in the number of active cases, resolved cases, and deaths?

# D. Summary Table

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Summary statistics for key numerical variables
summary_stats <- summary(data[, c("ACTIVE_CASES", "RESOLVED_CASES", "DEATHS")])
summary_stats

# Visual exploration - Relationships between variables
# Scatterplot matrix to visualize pairwise relationships
pairs(data[, c("ACTIVE_CASES", "RESOLVED_CASES", "DEATHS")])

# Boxplot to visualize distribution of Active Cases, Resolved Cases, and Deaths
par(mfrow = c(1, 3))  # Set plot layout to 1 row and 3 columns
boxplot(data$ACTIVE_CASES, main = data$ACTIVE_CASES)
boxplot(data$RESOLVED_CASES, main = data$RESOLVED_CASES)
boxplot(data$DEATHS, main = data$DEATHS)
```

# Perform Hypothesis Testing and Calculate Confidence Intervals

## Calculate A Confidence Interval for The "Average of The ACTIVE_CASES Variable"
```{r, echo=FALSE, warning=FALSE, message=FALSE}
active_sample = data$ACTIVE_CASES
active_print <- t.test(active_sample)

print(active_print)
```

## Calculate A Confidence Interval for The "Average of The RESOLVED_CASES Variable"
```{r, echo=FALSE, warning=FALSE, message=FALSE}
resolved_sample = data$RESOLVED_CASES
t.test(resolved_sample)
```

## Calculate A Confidence Interval for The "Average of The DEATHS Variable"
```{r, echo=FALSE, warning=FALSE, message=FALSE}
deaths_sample = data$DEATHS
t.test(deaths_sample)
```

## Calculate A Confidence Interval for The "Proportion of ACTIVE_CASES above 448"

We are creating a Bernoulli variable named "value1_CI", where value1_CI = 1 if ACTIVE_CASES>448; value1_CI = 0 if ACTIVE_CASES<=448
```{r, echo=FALSE, warning=FALSE, message=FALSE}
data1_CI = data %>% mutate(value1_CI = as.numeric(ACTIVE_CASES>448, na.rm=T))

prop.test(x = sum(data1_CI$value1_CI), n = length(data1_CI$value1_CI))
```

## Calculate A Confidence Interval for The "Proportion of RESOLVED_CASES above 18000"

We are creating a Bernoulli variable named "value2_CI", where value2_CI = 1 if RESOLVED_CASES>18000; value2_CI = 0 if RESOLVED_CASES<18000
```{r, echo=FALSE, warning=FALSE, message=FALSE}
data2_CI = data %>% mutate(value2_CI = as.numeric(RESOLVED_CASES>18000, na.rm=T))

prop.test(x = sum(data2_CI$value2_CI), n = length(data2_CI$value2_CI))
```

## Calculate A Confidence Interval for The "Proportion of DEATHS above 245"

We are creating a Bernoulli variable named "value3_CI", where value3_CI = 1 if DEATHS>245; value3_CI = 0 if DEATHS<245
```{r, echo=FALSE, warning=FALSE, message=FALSE}
data3_CI = data %>% mutate(value3_CI = as.numeric(DEATHS>245, na.rm=T))

prop.test(x = sum(data3_CI$value3_CI), n = length(data3_CI$value3_CI))
```